Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-09 paper.tex

\begin{figure}
    \begin{tabular}{c}
\resizebox{7cm}{!}{\input{diagrams/data-model.tex}} \\
\includegraphics[width=7cm]{figures/data-scaling}
\end{tabular}
\caption{Chunked compressed storage of the genotype matrix using Zarr.
(A) Schematic of data stored under the \texttt{vcfzarr} specification.
(B) Compression performance compared with BCF, savvy and genozip
based on simulated data (see text for details).
\label{fig-data-storage}}
\end{figure}

We propose the \texttt{vcfzarr} specification which maps the 
VCF data model into a columnar layout using the
widely-used Zarr format (Fig~\ref{fig-data-storage}A). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the \texttt{vcfzarr} specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
Note that this approach to storing higher dimensional 
sets Zarr apart from more general columnar formats 
such as Parquet [TODO more details and citations?].
See the Methods for more detail on Zarr and the \texttt{vcfzarr}
specification.

Remarkably, the simple approach of compressing
two dimensional chunks with the Zstandard 
compressor~\citep{collet2021rfc} and the bit-shuffle
filter from Blosc~\cite{alted2010modern} produces 
compression levels competitive with specialised methods
on the genotype matrix. In Fig~\ref{fig-data-storage}B
we show compression performance on up to a million
samples subset a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so provides 
a reasonable data point when comparing such methods.
% Could note here that this might represent a best 
% case scenario for specialised methods, but maybe 
% not worth bothering with.

>>>>

\begin{figure}
\resizebox{225pt}{!}{\input{diagrams/data-model.tex}}
\caption{Columnar storage of VCF data in chunked, compressed form using Zarr.
[TODO more detail, including brief overview of how data maps from 
VCF to vcfzarr in the examples]
\label{fig-data-model}}
\end{figure}

We propose the \texttt{vcfzarr} specification which maps the 
VCF data model into a columnar layout using the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the \texttt{vcfzarr} specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
Note that this approach to storing higher dimensional 
sets Zarr apart from more general columnar formats 
such as Parquet [TODO more details and citations?].
See the Methods for more detail on Zarr and the \texttt{vcfzarr}
specification.

\begin{figure}
\begin{center}
\includegraphics[]{figures/data-scaling}
\end{center}
\caption{Compression performance on simulated genotypes.
Comparison of total stored bytes for VCF data produced 
by subsets of a large simulation of French-Canadians.
Sizes for $10^6$ samples are shown on the right. Sizes 
for Savvy (21.25GiB) and Zarr (22.06GiB) are very similar.
\label{fig-data-storage}}
\end{figure}

Remarkably, the simple approach of compressing
two dimensional chunks with the Zstandard 
compressor~\citep{collet2021rfc} and the bit-shuffle
filter from Blosc~\cite{alted2010modern} produces 
compression levels competitive with specialised methods
on the genotype matrix. In Fig~\ref{fig-data-storage}
we show compression performance on up to a million
samples subset a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so provides 
a reasonable data point when comparing such methods.
% Could note here that this might represent a best 
% case scenario for specialised methods, but maybe 
% not worth bothering with.
